---
title: "mae_demo"
output: pdf_document
date: '2023-02-10'
---

```{r echo=TRUE}
library(MultiAssayExperiment)
library(genefilter)
library(gplots) 
library(mogsa)
library(tidyverse)
library(patchwork)
library(stats)
library(knitr)

# MOGSA Analysis
library(mogsa)

# plotting
library(UpSetR)
library(igraph)
library(ggplot2)
library(gplots)

# survival analysis
library(survival)
library(survminer)
library(SingleCellExperiment)
```


Explore two major analysis techniques, one is mogsa (multi omics data integrative clustering and gene set analysis), the other one is mofa (multi omics factor analysis)
```{r}
data(miniACC)

data <- as.data.frame(colData(miniACC))

# each row is an observation (patient)
bio_info <- mergeReplicates(as.data.frame(wideFormat(miniACC)))
bio_info <- drop_na(bio_info) # 43 patients remaining
```

# Check Baseline Characteristics
```{r}
dead <- data[data$vital_status==1, ]
survived <- data[data$vital_status==0, ]

Mean.IQR.by.trt <- function(y,trt,decp=1){
  groups <- sort(unique(trt))
  all <- quantile(y)
  g1 <- quantile(y[trt==groups[1]]) # indicates hormone == 1
  g2 <- quantile(y[trt==groups[2]]) # indicates hormone == 2
  
  result <- matrix(NA,1,3)
  colnames(result) <- c(groups,"Overall")
  # g1[2]: 25% quantile, g1[3]: median (50% quantile), g1[4]: 75% quantile
  result[1,1] <- paste0(round(g1[3],decp)," (",round(g1[2],decp),", ",round(g1[4],decp),")")
  result[1,2] <- paste0(round(g2[3],decp)," (",round(g2[2],decp),", ",round(g2[4],decp),")")
  result[1,3] <- paste0(round(all[3],decp)," (",round(all[2],decp),", ",round(all[4],decp),")")
  return(result)
}

N.prct.by.trt <- function(x,trt,decp=1){
  groups <- sort(unique(trt))
  x.levels <- sort(unique(x))
  p <- length(x.levels)
  n <- length(x)
  n1 <- length(x[trt==groups[1]]) # total number of patients have no hormone
  n2 <- length(x[trt==groups[2]]) # total number of patients have hormone
  
  result <- matrix(NA,p,3)
  colnames(result) <- c(groups,"Overall")
  rownames(result) <- x.levels
  
  for (i in 1:p){
    # calculate the total number of patients with non-hormone on each level
    n1i <- sum(x[trt==groups[1]]==x.levels[i])
    # calculate the total number of patients with hormone on each level
    n2i <- sum(x[trt==groups[2]]==x.levels[i])
    ni <- sum(x==x.levels[i])
    
    
    result[i,1] <- paste0(n1i," (",round(n1i/n1*100,decp),"%)")
    result[i,2] <- paste0(n2i," (",round(n2i/n2*100,decp),"%)")
    result[i,3] <- paste0(ni," (",round(ni/n*100,decp),"%)")
  }
  
  
  return(result)
}


table1 <- rbind(
  Mean.IQR.by.trt(data$years_to_birth, data$vital_status), # age
  N.prct.by.trt(data$gender, data$vital_status), #gender
  N.prct.by.trt(data$pathologic_stage[complete.cases(data$pathologic_stage)], data$vital_status[complete.cases(data$pathologic_stage)]),
  Mean.IQR.by.trt(data$date_of_initial_pathologic_diagnosis, data$vital_status)
)

row.names(table1)<-c('Age', 'Female', 'Male', 'Stage I', 'Stage II', 'Stage III', 'Stage IV', 'Date of Initial Pathologic Diagnosis')

colnames(table1) <- c('Survived', 'Dead', 'Overall')

knitr::kable(table1,caption="Adrenocortical Carcinoma Baseline Characteristics")
```

# Survival Analysis
```{r}
data$y <- Surv(data$days_to_death, data$vital_status)

# Kaplan - Meier Curve
KMfit <- survfit(y ~ pathologic_stage, data = data)
ggsurvplot(
  KMfit, 
  cex.axis=1,
  cex.lab=1,
  cex.main=1, 
  lwd=2, 
  risk.table = F, 
  data = data,
  # x = 'Days to Death',
  # title = 'Kaplan Meier Curves Based on Pathologic Stages'
  ) +
  labs(
  x = 'Days to Death',
  title = 'Kaplan Meier Curves Based on Pathologic Stages'
  )

```


# Estimating correlation between RNA & Copy Number
```{r}
# For the gene with highest correlation to copy number, make a box plot of log2 expression against copy number.
# correlation between RNASeq2GeneNorm and copy number
subacc <- miniACC[, , c('RNASeq2GeneNorm', 'gistict')]

subacc.list <- assays(subacc)
# log transformation of the RNA-seq assay
# +1 avoids taking log of zeroes
subacc.list[[1]] <- log2(subacc.list[[1]] + 1)

corres <- cor(subacc.list[[1]], subacc.list[[2]])

hist(corres)
cat(which.max(diag(corres)))
# represent copy numer CHEK2
```

```{r}
df <- wideFormat(subacc["CHEK2", , ])
head(df)

boxplot(RNASeq2GeneNorm_CHEK2 ~ gistict_CHEK2,
        data=df, varwidth=TRUE,
        xlab="GISTIC Relative Copy Number Call",
        ylab="RNA-seq counts",
        main='CHEK2')
```
# Prepare for Mogsa
```{r}
# return those biological units (here are patients) that have measurements across all experiments
mae <- intersectColumns(miniACC[, , c('RNASeq2GeneNorm', 'gistict', 'Mutations')])
```


# Mogsa Analysis
```{r}
# Create support data
set.seed(123)
gene_names <- sample(rownames(mae[[1]]))

# every 10 genes belong to one cluster hypothetically, the last group contains only 8 genes
cluster.names <- c()
for (i in 1:20) {
  n <- paste0('gene.cluster.', i)
  cluster.names <- append(cluster.names, n)
}

# Create clustered gene sets, a list containing 10 vectors
gene_set <- list()
for (i in 1:20) {
  vec <- c()
  if (length(gene_names) >= 10){
    vec <- list(gene_names[1:10])
  } else {
    vec <- list(gene_names[1:length(gene_names)])
  }
  gene_names <- gene_names[11:length(gene_names)]
  
  gene_set <- append(gene_set, vec)
}

# main data for mogsa
gene <- list()
gene$rna <- assay(mae[[1]])
gene$gistict <- assay(mae[[2]])
gene$mutation <- assay(mae[[3]])

# create a supp matrix, row is gene names, column is the gene set
# 1 means this gene belongs to this gene set
# Create 3 different matrix, where the number of rows of each matrix corresponds to the number of rows of each assay info in mae
container <- list()

for (i in 1:3) {
  mat <- matrix(0, nrow=nrow(gene[[i]]), ncol=length(cluster.names))
  colnames(mat) <- cluster.names
  rownames(mat) <- rownames(gene[[i]])
  for (r in 1:nrow(mat)) {
    for (c in 1:ncol(mat)) {
      if (rownames(mat)[r] %in% gene_set[[c]]) {
        mat[r, c] <- 1
      }
    }
  }
  container <- append(container, list(mat))
}

supp <- list()
supp$rna <- container[[1]]
supp$gistict <- container[[2]]
supp$mutation <- container[[3]]
mgsa1 <- mogsa(x=gene, sup=supp, nf=10, proc.row = "center_ssq1", w.data = "inertia", statis = TRUE)
```

```{r}
# The variance of each principal components (PC)
eigs <- getmgsa(mgsa1, "partial.eig") # get partial "eigenvalue" for separate data
barplot(as.matrix(eigs), legend.text = rownames(eigs))
# not sure why there are 75 principal components
```


```{r}
scores <- getmgsa(mgsa1, "score") # get the gene set sample pathway scores
heatmap.2(scores, trace = "none", scale='r', Rowv=NA, Colv=T,margins=c(12,12))

# find the cluster with the maxmimal score
max.score <- names(which(rowSums(scores)==max(rowSums(scores))))
max.score

# cluster 4 has the highest gene set pathway scores
```
(Remember to Revise)It is also interesting to look into more detailed information for a specific gene set. For example, which dataset(s) contribute most to the high or low gene set score of a gene set?
And which genes are most important in defining the gene set score for a gene set? The former
question could be answered by the gene set score decomposition; the later question could
be solve by the gene influential score. 
```{r}
### Part 1 Find most significant gene set(cluster) by estimating the #of significant p-values ###
p.mat <- getmgsa(mgsa1, "p.val") # get p value matrix
# select gene sets with most signficant GSS scores.
# by summing up the number of each gene's p value < 0.01 across all patients
top.gs <- sort(rowSums(p.mat < 0.01), decreasing = TRUE) # total length is 20 bc of there are only 20 clusters
top.gs.name <- names(top.gs)

# gene.cluster.16 is the most significant one
# gene_set[[16]] "PXN"    "MAPK1"  "XBP1"   "RAB11B" "EGFR"   "PGR"    "BCL2"   "SMAD4"  "MAPK14" "TYMS"
gs1 <- top.gs.name[1] # select the most significant gene set
gs1


### Part 2 Find the genes are positive correlated with gene set scores ###

# gistict (copy number) & rna dataset has slightly higher influential scores
# The expression of genes with high positive GIS more likely to have a good positive correlation with the gene set score.
gis1 <- GIS(mgsa1, gs1)
head(gis1) # first 6 genes highly positively correlated with gene set scores
```


# Multi-Omics Factor Analysis
```{r}
library(MOFA2)
library(basilisk)

# Use both mogsa and mofa2 to analysis mae
mofa_obj <- create_mofa(mae)

# Data Inspection
# plot_data_overview(mofa_obj)
print(mofa_obj) # rownames as feature (colnames are patient id's), 92 samples per group
```

```{r echo=TRUE}
# Define Data option
data_opts <- get_default_data_options(mofa_obj)
# Normalization works worse when presenting heatmap though it reduces variations among factors
data_opts$center_groups <- FALSE
data_opts$scale_groups <- FALSE

# Define Model option, dont change the default option unless familiar with the underlying mathmatical models
model_opts <- get_default_model_options(mofa_obj)
# head(model_opts)

# Define Training option
train_opts <- get_default_training_options(mofa_obj)
# head(train_opts)

# Prepare mofa object
mofa_obj <- prepare_mofa(
  object = mofa_obj,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

# training
outfile = file.path(tempdir(),"model.hdf5")
mofa_obj.trained <- run_mofa(mofa_obj, outfile, use_basilisk = TRUE)
plot_variance_explained(mofa_obj.trained)
```
Observation, factor 1 is the most variational factor among gistict assay
```{r}
v <- calculate_variance_explained(mofa_obj.trained)
v_gistict_max <- max(as.data.frame(v$r2_per_factor)[, 2])
max.percent <- v_gistict_max/(as.data.frame(v$r2_total)[2,])
paste0(round(max.percent * 100, 2), '%') # factor 1 accounts for 19.02 of total R^2 among all gistict factors
```


```{r}
# Coefficient of Determination (R^2)
v$r2_total

# by checking the coefficient of determination, we observe that gistict dataset fit the MOFA model the best
```
```{r}
# Violin plot of first three factors stratified by gender
# thinner part represents lower probability

plot_factor(mofa_obj.trained, 
  factor = 1:3,
  color_by = 'gender',
  add_violin = T,
  violin_alpha = 0.25,
  dodge = T
)
```
```{r}
# Distribution and scatter plots of first three factors
plot_factors(
  mofa_obj.trained,
  factors = 1:3,
  color_by = 'SCNA.cluster'
  )
```
From previous variance plot, we would like to investigate the factor1 in gistict, by checking the weights of each feature and their correlations.

Top 5 weights
```{r}
plot_top_weights(mofa_obj.trained,
  view = "gistict",
  factor = 1,
  nfeatures = 5
)

```
Heatmap, we seek to present the heat map for which factor has highest R^2
RNASeq2GeneNorm
```{r}
# RNASeq2GeneNorm
v_factor <- as.data.frame(v$r2_per_factor)
which(v_factor[, 1] == max(v_factor[, 1]))

# columns are patient id
# also report some deep orange grids
plot_data_heatmap(mofa_obj.trained,
  view = "RNASeq2GeneNorm",
  factor = which(v_factor[, 1] == max(v_factor[, 1])),            
  features = 10,          

  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, 
  cluster_cols = TRUE,
  show_rownames = TRUE, 
  show_colnames = TRUE
)
```
Gistict, copy number
```{r}
# gistict
v_factor <- as.data.frame(v$r2_per_factor)
which(v_factor[, 2] == max(v_factor[, 2]))

# columns are patient id
# also report some deep orange grids
plot_data_heatmap(mofa_obj.trained,
  view = "gistict",
  factor = which(v_factor[, 2] == max(v_factor[, 2])),            
  features = 10,          

  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, 
  cluster_cols = TRUE,
  show_rownames = TRUE, 
  show_colnames = TRUE
)
```



Same way to check the mutation, find the factor with highest R^2 first
Mutations
```{r}
# Inspecting the highly correlated genes in Mutations
which(v_factor[, 3] == max(v_factor[, 3]))

plot_data_heatmap(mofa_obj.trained,
  view = "Mutations",
  factor = which(v_factor[, 3] == max(v_factor[, 3])),            
  features = 10,          

  # extra arguments that are passed to the `pheatmap` function
  cluster_rows = TRUE, 
  cluster_cols = TRUE,
  show_rownames = TRUE, 
  show_colnames = TRUE
)
```
